---
title: "FINAL RESULTS REPORT"
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: true
css: "config/report.styles.css"
format:
  html:
    page-layout: full
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    toc: true
    toc-expand: true
    toc-location: right
    toc-title: Summary
    number-sections: true
editor: source
params:
  heatmap: ""
  cell_counts: ""
  violin_counts: ""
  violin_features: ""
  HGVs_plot: ""
  elbowplot: ""
  feature_plot_path: ""
  de_summary: ""
  gsea_table: ""
  annotated_genes: ""
  legends: ""
  reductions: ""
  resolution: 0.0
  algorithm: 0
  markers_logfc_threshold: 0.0
  markers_min_pct: 0.0
  markers_test: ""
  markers_heatmap: 0
  dim_plot_path: ""
  gsea_pval: 0.0
  gsea_fdr: 0.0
  gsea_padj_method: ""
  gsea_mingenes: 0 
  de_fdr: 0.0
  de_fc: 0.0
  sc_min_cells: 0
  sc_min_features: 0
  sc_scale_factor: 0
  sc_normalization_method: ""
  sc_selection_method: ""
  sc_nfeatures: 0
output-file: report.html
---


```{r libraries-load, include = F,warning = F}
library(DT)
library(tidyverse)
```

## Study inspection
The distribution of cells across samples is shown in @tbl-summary. It also reports the total number of cells. 
Counts correspond to filtered data, retaining cells with a minimum number of genes per cell (`r params$sc_min_features`) 
and genes expressed in at least a minimum number of cells (`r params$sc_min_cells`).

```{r tbl-summary, echo = F,message = F, warning = F}
#| tbl-cap: Distribution of cells across samples and total cell counts after quality filtering.

table <- read.csv(params$cell_counts)

DT::datatable(table)
```

Violin plots allow the visualization of the distribution of counts (@fig-violin_counts) and the number of detected 
features or genes (@fig-violin_features) between cases and controls. These plots facilitate the identification of variability, skewness, and 
potential outliers across cells. The distributions shown correspond to data normalized using the **_`r params$sc_normalization_method`_** 
method with a scale factor of `r params$sc_scale_factor`, enabling comparison across cells by correcting for differences in sequencing depth.

::: {.panel-tabset}
## N. Counts

![Violin plot showing the distribution of counts after normalization and scale dara. 
The plot highlights variability and potential outliers across cells.](`r params$violin_counts`){#fig-violin_counts, width=60%}

## N. Features
![Violin plot showing the distribution of the number of detected features (genes) after normalization and scale data. 
This visualization facilitates the assessment of gene detection variability across cells.](`r params$violin_features`){#fig-violin_features, width=60%}
:::


Highly Variable Genes (HVGs) capture the majority of the variability present in the data and are therefore used for downstream analyses.
 These genes are selected to emphasize biologically meaningful variation while reducing technical noise. The top `r params$sc_nfeatures` 
 features are identified using the **_`r params$sc_selection_method`_** method to represent the HVGs. The remaining genes are expected 
 to exhibit lower variability and are retained in the background, as illustrated in @fig-HGVs_plot.
 
![Identification of highly variable genes (HVGs) highlighted in red, while the remaining genes are shown in black and
 represent features with lower variability.](`r params$HGVs_plot`){#fig-HGVs_plot, width=60%}


## Clustering using Principal component analysis (PCA)


```{r get_algorithm, include = F}
n = params$algorithm
algorithm_name = ""

if (n == 1) {
  algorithm_name = "Louvain algorithm"
} else if (n == 2) {
  algorithm_name = "Louvain algorithm with multilevel refinement"
} else if (n == 3) {
  algorithm_name = "SLM algorithm"
} else if (n == 4) {
  algorithm_name = "Leiden algorithm"
} else {
  algorithm_name = "Unknown algorithm"
}

```


Principal component analysis (PCA) was performed on scaled data to reduce dimensionality and 
capture the main sources of variation. The optimal number of PCs used for downstream analyses was determined using an elbow plot, 
which identifies the point at which the marginal gain in explained variance decreases substantially (@fig-elbowplot).

Based on this criterion, the first principal components up to the elbow point were selected for neighborhood graph construction. 
These components were then used to compute a shared nearest neighbor (SNN) graph, followed by graph-based clustering using the 
**_`r algorithm_name`_** (algorithm = `r params$algorithm`) with a resolution of `r params$sc_resolution`, allowing identification 
of transcriptionally distinct cell populations.

![Elbow plot showing the standard deviation of each principal component (PC) obtained from PCA. The elbow point indicates the number of PCs retained for downstream analyses, 
beyond which additional components contribute marginally to explaining variance.](`r params$elbowplot`){#fig-elbowplot, width=60%}

Marker genes were identified for each cluster to characterize and assign cell identities. Top genes were selected by  
using the **_`r params$markers_test`_** test, applying a minimum log fold-change threshold of `r params$markers_logfc_threshold` and 
requiring genes to be expressed in at least `r params$markers_min_pct` of cells within a cluster. The top marker genes annotated regards cell identities 
for each cluster are reported in @tbl-genes.


```{r tbl-genes, echo = F,message = F, warning = F}
#| tbl-cap: Top marker genes identified for each cluster. These genes were used to represent cluster-specific transcriptional profiles and support cell identity annotation.

table <- read.csv(params$annotated_genes) %>%
  rename(
    Cluster = cluster,
    Marker = marker,
    "Associated Cells Types" = cells
  ) 

DT::datatable(table, rownames = FALSE)
```

### Heatmap

To inspect the marker genes associated with each cluster, we visualized the expression of the top `r params$markers_heatmap` marker genes
 across clusters using a heatmap (@fig-heatmap).

![Heatmap showing the expression patterns of the top marker genes across clusters. Lighter colors indicate higher gene expression levels, 
while darker colors represent lower expression, facilitating comparison of cluster-specific transcriptional profiles.](`r params$heatmap`){#fig-heatmap}


## Dimensionality reduction

To explore the global structure of the data and assess cluster separation, dimensionality reduction techniques were applied and
 visualized using two-dimensional embeddings. For each reduction method, cells are displayed according to different metadata 
 annotations, allowing comparison of cell distributions and evaluation of potential batch effects or biological patterns.

In addition, feature plots are shown for each reduction, illustrating the expression of selected marker genes across the 
low-dimensional space. These visualizations facilitate the interpretation of cluster identities and the relationship between 
transcriptional programs and the inferred cell populations.


```{r dim_plots, results="asis", echo = F}

a_vals <- unlist(strsplit(params$reductions,";"))
b_vals <- unlist(strsplit(params$legends,";"))

for (a in a_vals) {

  cat("## Reduction:", a, "\n\n")
  cat("::: {.panel-tabset}\n\n")

  for (b in b_vals) {

    cat("### ", b, "\n\n")

    path <- params$dim_plot_path
    path <- gsub("\\(reduction\\)", a, path)
    path <- gsub("\\(legend\\)", b, path)

    

    cat("![Two-dimensional embedding of cells obtained using the indicated dimensionality reduction method. Cells are colored according to the specified metadata annotation.](", path, "){width=80%}\n\n")
  }
  cat(":::\n\n")

  cat("### Features\n\n")
  path2 <- params$feature_plot_path
  path2 <- gsub("\\(reduction\\)", a, path2)
  cat("![Feature plot showing the expression of selected marker genes projected onto the low-dimensional embedding. Color intensity reflects gene expression levels.](", path2, "){width=80%}\n\n")
  
}
```


## Differencial expression of genes

Differencial expression in genes for each cluster was summarized @tbl-de. Differences in cell distribution 
between case and control samples were also evaluated at the cluster 
level to identify changes in cell population abundance associated with the condition of interest and compare to change 
regards genes expression. For each cluster, the proportion of cells per sample was calculated and summarized as mean 
 percentage ± standard deviation for case and control groups. Statistical comparisons between groups were performed using 
 both t-test and Wilcoxon rank-sum test on cell percentage distributions. In addition, the log2 fold change of cell proportions between 
 case and control groups was computed to quantify the direction and magnitude of abundance differences. 

```{r tbl-de, echo = F,message = F,warning = F}
#| tbl-cap: Differential expression and summary of cell distribution differences between case and control samples across clusters. 

table <- read.csv(params$de_summary) %>%
  rename(
    "% Case cells" = Case_._cells,
    "% Control cells" = Control_._cells,
    "P.value (t.test)" = P.value..t.test.,
    "P.value (wilcox)" = P.value..wilcox.   
  ) 

DT::datatable(table, rownames = FALSE)
```

## Gene Set Enrichment Analysis (GSEA)

Gene Set Enrichment Analysis (GSEA) was performed for each cluster to evaluate whether a 
predefined gene signature was significantly enriched among differentially expressed genes. 
Only clusters with at least `r params$gsea_mingenes` differentially expressed genes were 
considered for GSEA. Filters was evaluated using a nominal p-value 
threshold of `r params$gsea_pval` and a false discovery rate (FDR) cutoff of `r params$gsea_fdr`,
 with p-values adjusted according to the **_`r params$gsea_padj_method`_** method (@tbl-gsea).

```{r tbl-gsea, echo = F,message = F,warning = F}
#| tbl-cap: Gene Set Enrichment Analysis (GSEA) results for each cluster based on a predefined gene signature.

table <- read.csv(params$gsea_table)  %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  select(!cluster)

DT::datatable(table, rownames = FALSE)
```
