---
title: "FINAL RESULTS REPORT"
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: true
css: "config/report.styles.css"
format:
  html:
    page-layout: full
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    toc: true
    toc-expand: true
    toc-location: right
    toc-title: Summary
    number-sections: true
editor: source
params:
  heatmap: ""
  cell_counts: ""
  feature_plot: ""
  de_summary: ""
  gsea_table: ""
  annotated_genes: ""
  legends: ""
  reductions: ""
  dim_plot_path: ""
  gsea_pval: 0.0
  gsea_fdr: 0.0
  gsea_padj_method: ""
  gsea_mingenes: 0 
  de_fdr: 0.0
  de_fc: 0.0
  sc_min_cells: 0
  sc_min_features: 0
  sc_scale_factor: 0
  sc_normalization_method: ""
  sc_selection_method: ""
  sc_nfeatures: 0
output-file: report.html
---


```{r libraries-load, include = F,warning = F}
library(DT)
library(tidyverse)
```

## Summary table

```{r tbl-summary, echo = F,message = F, warning = F}
#| tbl-cap: Target dataset summary

table <- read.csv(params$cell_counts)

DT::datatable(table)
```

## Annotated Markers 

```{r tbl-genes, echo = F,message = F, warning = F}
#| tbl-cap: Target dataset summary

table <- read.csv(params$annotated_genes) %>%
  rename(
    Cluster = cluster,
    Marker = marker,
    "Associated Cells Types" = cells
  ) 

DT::datatable(table, rownames = FALSE)
```

```{r DE, echo = F,message = F,warning = F}
#| tbl-cap: Target dataset summary

table <- read.csv(params$de_summary) %>%
  rename(
    "% Case cells" = Case_._cells,
    "% Control cells" = Control_._cells,
    "P.value (t.test)" = P.value..t.test.,
    "P.value (wilcox)" = P.value..wilcox.
    
  ) 

DT::datatable(table, rownames = FALSE)
```

```{r GSEA, echo = F,message = F,warning = F}
#| tbl-cap: Target dataset summary

table <- read.csv(params$gsea_table)  %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  select(!cluster)

DT::datatable(table, rownames = FALSE)
```

```{r dim_plots, results="asis", echo = F}

a_vals <- unlist(strsplit(params$reductions,";"))
b_vals <- unlist(strsplit(params$legends,";"))

for (a in a_vals) {

  cat("## Reduction:", a, "\n\n")
  cat("::: {.panel-tabset}\n\n")

  for (b in b_vals) {

    cat("### ", b, "\n\n")

    path <- params$dim_plot_path
    path <- gsub("\\(reduction\\)", a, path)
    path <- gsub("\\(legend\\)", b, path)

    

    cat("![path](", path, "){width=80%}\n\n")
  }

  cat(":::\n\n")
}
```

## Plots

### Heatmap

The lineage admixture for each sample has been estimated
using [Freyja](https://github.com/andersen-lab/Freyja) (@fig-heatmap).

![Estimated lineage admixture of each sample.
Samples in the X-axis are ordered chronologically, from more ancient to newer.](`r params$heatmap`){#fig-heatmap}

### Features plot

The lineage admixture for each sample has been estimated
using (@fig-feature_plot).

![Estimated lineage admixture of each sample.
Samples in the X-axis are ordered chronologically, from more ancient to newer.](`r params$feature_plot`){#fig-feature_plot}
